---
title: "SigFun"
author:
- name: Wei-Chung Cheng
  affiliation: China Medical University
output:
  BiocStyle::html_document:
    toc_float: true
abstract: >
    SigFun is an R package designed to streamline the analysis of 
    transcriptomic data in relation to specific gene signatures. It provides an 
    automated workflow for analyzing gene signatures and their functional 
    implications in transcriptomic datasets. The package integrates Gene Set 
    Enrichment Analysis (GSEA) with visualization tools to help researchers 
    understand the biological pathways and processes associated with their gene 
    signatures of interest.

vignette: >
  %\VignetteIndexEntry{SigFun}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo=FALSE, results='asis'} 
BiocStyle::markdown()
```

# Table of contents
[Introduction](#intro)
[Installation](#install)
[Input data demonstration](#input)
[SigFun one-click command](one-click)
[Step 1. sigCor](#sigcor)
[Step 2. sig2GSEA](#gsea)
[Step 3. plot_bar](#plot_bar)
[Step 4. plot_heat](#plot_heat)
[Session info](#info)

# Introduction {#intro}
    SigFun is an R package designed to streamline the analysis of 
    transcriptomic data in relation to specific gene signatures. It provides an 
    automated workflow for analyzing gene signatures and their functional 
    implications in transcriptomic datasets. The package integrates Gene Set 
    Enrichment Analysis (GSEA) with visualization tools to help researchers 
    understand the biological pathways and processes associated with their gene 
    signatures of interest.

# Installation  {#install}
Here is the procedures of running the **SigFun** package on your system. 
We assume that you have already installed the R program (see the R project at 
[http://www.r-project.org](http://www.r-project.org)  and are familiar with it. 
You need to have R 4.4.0 or a later version installed for running SigFun.

Our package is available at the Bioconductor repository 
[http://www.bioconductor.org](http://www.bioconductor.org). 
To install our package, first, you need to install the core Bioconductor 
packages. If you have already installed the Bioconductor packages on your 
system, you can skip the following step below.

```{r install_Bioconductor, eval=FALSE}
if (!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install()
```

Once the core Bioconductor packages have been installed, we can begin the 
installation of the SigFun package.

```{r install_package, eval=FALSE}
if (!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install("SigFun")
``` 

After conducting the above step, now you can load in our package and start 
using it!

```{r load, message=FALSE}
library(SigFun)
library(BiocStyle)
library(S4Vectors)
```

```{r load_package, message=FALSE, echo=FALSE}
library(dplyr)
library(utils)
```

# Input data demonstration {#input}
Load the demo data and investigate how to prepare the input
```{r abundance}
rm(list=ls())
data("demo")
slotNames(SE_data)
```

# Input preparation
Prepare you data as the demo data
```{r char_table}
# Transcriptome dataset
GENE_MAT <- as.data.frame(SummarizedExperiment::assay(SE_data))
# Signature score
SIG_MAT <- as.data.frame(SummarizedExperiment::rowData(SE_data))
# gene symbol mapping table (ensemble ID and official gene symbol lookup table)
mapping <- as.data.frame(SummarizedExperiment::colData(SE_data))
```

Construct SummarizedExperiment object from your data 
```{r build_se}
assay.data <- GENE_MAT
rowData.data <- SIG_MAT
colData.data <- mapping

SE_data <- SummarizedExperiment::SummarizedExperiment(
    assays=list(abundance=as.matrix(assay.data)),
    rowData=S4Vectors::DataFrame(rowData.data, 
    row.names=rowData.data$sample_id), colData=colData.data)
```

# SigFun one-click command {#one-click}
SigFun accepts the following input parameters:

- `SE_data`: A SummarizedExperiment object containing your expression data
- `ranking.method`: Method for ranking genes ("stat" or other supported methods)
- `species`: Organism of study (e.g., "human")
- `cor.method`: Correlation method for analysis (e.g., "spearman")
- `pathways.all`: Collection of pathways for enrichment analysis
- `output_path`: Directory for saving results
- `topN`: Number of top pathways to return
- `Z.transform`: Logical indicating whether to perform Z-transformation
- `significat_type`: Type of significance measure ("pval" or alternatives)
- `strings`: Vector of pathway databases to query (e.g., "GOBP")

```{r SigFun}
# Set output directory
output_path <- tempdir()
dir.create(output_path,TRUE)
# Run SigFun analysis
res <- sig2Fun(
    SE_data=SE_data,
    ranking.method="stat",
    species="human",
    cor.method="spearman",
    pathways.all=pathways.all,
    output_path=output_path,
    topN=10,
    Z.transform=FALSE,
    significat_type="pval",
    strings=c("KEGG"),
    plot_out=TRUE
)
```

# Step 1. SigCor: run gene correlation with signature patterns {#sigcor}
```{r SigCor}
ranking.method="stat"
species="human"
cor.method="spearman"
topN=10
Z.transform=FALSE
significat_type="pval"
strings=c("KEGG")
SE_data.cor <- sigCor(SE_data=SE_data, cor.method=cor.method,
                            output_path=output_path, Z.transform=Z.transform)
```

# Step 2. Sig2GSEA: Run fGSEA using gene ranking values load corRES {#gsea}

```{r Sig2GSEA}
SE_data.fgsea <- sig2GSEA(SE_data.cor=SE_data.cor,
                        ranking.method=ranking.method,
                        output_path=output_path, 
                        pathways.all=pathways.all)
```

# Step 3. plot_bar: barplot for visualization {#plot_bar}

```{r plot.bar}
barplots <- plot_bar(SE_data.fgsea=SE_data.fgsea, 
                        output_path=output_path,
                        topN=topN, significat_type=significat_type, 
                        strings=strings)
```

# Step 4. plot_heat: heatmap for visualization {#plot_heat}

```{r heatmap}
heatmap <- plot_heat(SE_data.fgsea=SE_data.fgsea, 
                        output_path=output_path,
                        strings=strings, significat_type=significat_type, 
                        topN=topN, pathways.all=pathways.all, 
                        ranking.method=ranking.method)
```

# Session info {#info}
```{r sessionInfo, echo=FALSE}
sessionInfo()
```
